Undecidability and the structure of the Turing degrees

### 2018-08-13 paper.tex
Turing machine def

\section{Turing Machines}
Upon consideration of different types of problems, it quickly becomes clear that
some are much more difficult than others. For example, when considering words
over the alphabet $\{a,b,c\}$, it seems much easier to determine whether a word
ends in an $a$ than whether a word is a palindrome. It seems even harder still
to determine whether the word is of the form $a^nb^nc^n$. In order to formalize
these notions of difficulty, we need to build abstract models of computation,
and then test their ability to decide such questions. We observe that a machine
that can solve the first problem only needs to ``remember'' the same amount of
information no matter how long the input is: the last letter of the word. In
contrast, the amount of memory that the second and third problems require is
dependent on the size of the input, because the first half of the palindrome or
$n$, respectively, can be arbitrarily large. The difference between the second
and third problems is more subtle: the second problem can be solved moving only
one direction in memory, while the third problem requires the ability to ``look
back'' in memory.
\par
Based on these differences, we can begin to construct different models of
computation that have just enough power to solve each problem. The first problem
can be solved by a \emph{\ac{FSA}}, a collection of a finite
number of states, a transition function, and a set of accepting states. The
\ac{FSA} accepts the input if the sequential application of each element to the
input finishes in an accepting state, and rejects otherwise. \aclp{FSA} are often represented by charts as in Figure~\ref{fig:fsa}, a
representation of an \ac{FSA} that decides the first problem.
\begin{figure}[h]
  \caption{An \ac{FSA} that determines whether a string ends in
    `$a$'.}
  \label{fig:fsa}
  \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state,accepting] (q_1) [right=of q_0]{$q_1$};
  \path[->]
  (q_0) edge [loop above] node {b,c} ()
  edge [bend left] node  {a} (q_1)
  (q_1) edge [bend left] node {b,c} (q_0)
  edge [loop above] node {a} ();
\end{tikzpicture}
\end{figure}
\par
The second problem can be solved by a machine called a \emph{\ac{PDA}}, which is essentially a finite state automaton with the addition of
a stack. Rather than transforming the current state and the input into a new
state, as with a finite state automata, the transition function of a \ac{PDA} also considers the stack, which it can pop and push from. Within the
collection of \acp{PDA}, there is another division between
\emph{\acp{DPDA}} and \emph{Nondeterministic \aclp{PDA}}. For \acp{DPDA}, the
transition function only outputs one move, rather than a set of
moves. Nondeterministic \acp{PDA} are able to recognize more languages than
\acp{DPDA}. Figure~\ref{fig:pda} is a graphical representation of a \ac{PDA}
that decides palindromes. In this representation, $\epsilon$ is an empty input,
and  is a special symbol that denotes the bottom of the stack. The moves are
represented by a double of input and stack operation.

\begin{figure}[h]
  \label{fig:pda}
  \caption{A \acl{PDA} that decides palindromes over the alphabet ${a,b}$}
  \begin{tikzpicture}[shorten >=1pt,node distance=3cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state] (q_1) [right=of q_0]{$q_1$};
  \node[state] (q_2) [right=of q_1]{$q_2$};
  \node[state,accepting] (q_3) [right=of q_2]{$q_3$};
  \path[->]
  (q_0) edge node {$\epsilon,\epsilon\to\$$} (q_1)
  (q_1) edge [loop above] node[text width=1.5cm,align=center] {$0,\epsilon\to 0$
    $1,\epsilon\to 1$} ()
  edge node {$\epsilon,\epsilon\to\epsilon$} (q_2)
  (q_2) edge [loop above] node[text width=1.5cm,align=center] {$0,0\to\epsilon$
    $1,1\to\epsilon$} ()
  edge node {$\epsilon,\$\to\epsilon$} (q_3);  
  \end{tikzpicture}
\end{figure}
\par
Finally, we arrive at the \ac{TM}, devised by Alan Turing in his paper \emph{On
  Computable Numbers, With an Application To the Entscheidungsproblem}
~\cite{turing37_comput_number_with_applic_to_entsc}. \aclp{TM} offer an
unrestricted model of computation,\todo{Discussion of TMs}
\section{Undecidability}

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  \cite{turing37_comput_number_with_applic_to_entsc}
  \cite{sipser13:_introd_theor_comput}
\end{prob}
\begin{thm}[Rice's Theorem] \cite{kozen99_autom}
  % Proof by reduction to the halting problem?
\end{thm}
\subsection{Incompleteness}
\begin{thm}[G\"odel's Incompleteness Theorem] 
  % Kleene 1943? Also shown in hopcroft 1979
  \cite{kleene43_recur_predic_quant}
\end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable.
\cite{sipser13:_introd_theor_comput}
\cite{post44:_recur}

@@

\begin{defn}
  A Pushdown Automata is...
\end{defn}

@@

@@

\begin{prob}[Post's Problem]
  \cite{post44:_recur}
  % Marchenkov 1976
  % Independently by Friedberg 57, Mucnik 56/58 using priority method
  % Priority free solution Kucer 1986''
  % ``Natural'' solns don't exist
  \cite{Friedberg236}
\end{prob}

>>>>

\section{Models of Computation}
Upon consideration of different types of problems, it quickly becomes clear that
some are much more difficult than others. For example, when considering words
over the alphabet $\{a,b,c\}$, it seems much easier to determine whether a word
ends in an $a$ than whether a word is a palindrome. It seems even harder still
to determine whether the word is of the form $a^nb^nc^n$. In order to formalize
these notions of difficulty, we need to build abstract models of computation,
and then test their ability to decide such questions. We observe that a machine
that can solve the first problem only needs to ``remember'' the same amount of
information no matter how long the input is: the last letter of the word. In
contrast, the amount of memory that the second and third problems require is
dependent on the size of the input, because the first half of the palindrome or
$n$, respectively, can be arbitrarily large. The difference between the second
and third problems is more subtle: the second problem can be solved moving only
one direction in memory, while the third problem requires the ability to ``look
back'' in memory.
\par
Based on these differences, we can begin to construct different models of
computation that have just enough power to solve each problem. The first problem
can be solved by a \emph{\ac{FSA}}, a collection of a finite
number of states, a transition function, and a set of accepting states. The
\ac{FSA} accepts the input if the sequential application of each element to the
input finishes in an accepting state, and rejects otherwise. \aclp{FSA} are often represented by charts as in Figure~\ref{fig:fsa}, a
representation of an \ac{FSA} that decides the first problem.
\begin{figure}[h]
  \caption{An \ac{FSA} that determines whether a string ends in
    `$a$'.}
  \label{fig:fsa}
  \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state,accepting] (q_1) [right=of q_0]{$q_1$};
  \path[->]
  (q_0) edge [loop above] node {b,c} ()
  edge [bend left] node  {a} (q_1)
  (q_1) edge [bend left] node {b,c} (q_0)
  edge [loop above] node {a} ();
\end{tikzpicture}
\end{figure}
\par
The second problem can be solved by a machine called a \emph{\ac{PDA}}, which is essentially a finite state automaton with the addition of
a stack. Rather than transforming the current state and the input into a new
state, as with a finite state automata, the transition function of a \ac{PDA} also considers the stack, which it can pop and push from. Within the
collection of \acp{PDA}, there is another division between
\emph{\acp{DPDA}} and \emph{Nondeterministic \aclp{PDA}}. For \acp{DPDA}, the
transition function only outputs one move, rather than a set of
moves. Nondeterministic \acp{PDA} are able to recognize more languages than
\acp{DPDA}. Figure~\ref{fig:pda} is a graphical representation of a \ac{PDA}
that decides palindromes. In this representation, $\epsilon$ is an empty input,
and  is a special symbol that denotes the bottom of the stack. The moves are
represented by a double of input and stack operation.

\begin{figure}[h]
  \label{fig:pda}
  \caption{A \acl{PDA} that decides palindromes over the alphabet ${a,b}$}
  \begin{tikzpicture}[shorten >=1pt,node distance=3cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state] (q_1) [right=of q_0]{$q_1$};
  \node[state] (q_2) [right=of q_1]{$q_2$};
  \node[state,accepting] (q_3) [right=of q_2]{$q_3$};
  \path[->]
  (q_0) edge node {$\epsilon,\epsilon\to\$$} (q_1)
  (q_1) edge [loop above] node[text width=1.5cm,align=center] {$a,\epsilon\to a$
    $b,\epsilon\to b$} ()
  edge node {$\epsilon,\epsilon\to\epsilon$} (q_2)
  (q_2) edge [loop above] node[text width=1.5cm,align=center] {$a,a\to\epsilon$
    $b,b\to\epsilon$} ()
  edge node {$\epsilon,\$\to\epsilon$} (q_3);  
  \end{tikzpicture}
\end{figure}

\subsection{Turing Machines}
Finally, we arrive at the \ac{TM}, devised by Alan Turing in his paper \emph{On
  Computable Numbers, With an Application To the Entscheidungsproblem}
~\cite{turing37_comput_number_with_applic_to_entsc}. \aclp{TM} offer an
unrestricted model of computation, equivalent to the general purpose computers
that we are familiar with today. This equivalence is expressed by a pair of
concepts called \emph{Turing completeness} and \emph{Turing equivalence}. A
system of computation is Turing complete if it is able to simulate an arbitrary
Turing machine, and is Turing Equivalent if there is a Turing machine that can
simulate it. Essentially, Turing completeness is being \emph{at least} as
powerful as Turing machines, while Turing equivalence is being \emph{exactly} as
powerful as a Turing machine. Turing machines are themselves Turing complete, as are other
abstract models of computation such as the \emph{lambda calculus}. General
purpose programming languages are also considered Turing complete, because
although they are run on real machines with finite storage space, this space is
theoretically unbounded, and thus the expressive power of the languages are
unbounded even if any particular instance of a program run on an actual machine
would not be able to simulate all Turing machines. 
\par
A basic informal definition of a Turing machine is a collection of an infinite tape
of discrete cells,
a head that can move back and forth between cells on the tape as well as read and write
from the tape, a register that stores the current state of the machine out of a
finite set of possible states and starts in a start state, a set of accepting
states, and a transition function that at each step allows tells the machine
wether to write to the tape, move the head, or change states. There are
modifications to this definition which allow for multiple tapes or
nondeterminism, but they are all equivalent to the basic
definition. Additionally, a \ac{PDA} with two stacks is equivalent to a Turing
machine, as the stacks can be thought of as extending out in opposite
directions, forming an infinite tape.

\section{Undecidability}
In the previous section we were concerned primarily with problems that our
models of computation were able to solve. However, much as an \ac{FSA} is unable
to solve the problem of whether a string is a palindrome, there are problems
that are too difficult for even \aclp{TM} to solve. When a decision problem $A$
cannot be solved by a model $\mathcal{M}$, we say that $A$ is undecidable for
$\mathcal{M}$. When we just say that $A$ is undecidable, we mean that it is
undecidable for \acp{TM}. \par
It can be difficult to come up with obvious examples of problems that are
undecidable. After all, we usually talk about sets based on the patterns they
exhibit, and those patterns are typically simple enough to be enumerated by a
computer. A promising place to look for problems that a model cannot decide is
problems about the model itself, and the most famous undecidable problem is of
that type. The Halting Problem, introduced by Alan Turing in the same paper
where he introduced the Turing machine, asks whether it is possible to contsruct
a Turing machine that can determine whether a Turing machine run on a given
input will ever halt and return an
answer~\cite{turing37_comput_number_with_applic_to_entsc}. The proof is through
a method called \emph{diagonalizdation}, due to its similarity to Cantor's
diagonal argument.

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  Let $M$ be a Turing Machine and $i$ be an input. If $M$ is run on $i$, will it
  eventually halt?
  \begin{proof}[Proof of Undecidability]
    Suppose for the sake of contradiction that the halting problem is
    decidable. Then, there exists some machine
    $\mathcal{O}$ that can decide the halting problem. Then, we can construct a
    Turing machine $H$ that simulates $\mathcal{O}$ on its input $(M,i)$. In the
    case where $\mathcal{O}(M,i)$ accepts, $H$ enters into an infinite
    loop, whereas in the case where $\mathcal{O}(M,i)$ does not accept, $H$
    halts. Consider $H(H,\epsilon)$. If $H$ halts, then
    $\mathcal{O}(H,\epsilon)$ accepts, and so $H$ enters an infinite loop, and
    so it does not halt. But if $H$ does not halt, then
    $\mathcal{O}(H,\epsilon)$ does not accept, so $H$ halts. This is a
    contradiction, so our premise that the halting problem is decidable must be false.
  \end{proof}

  \cite{sipser13:_introd_theor_comput}
\end{prob}
\begin{thm}[Rice's Theorem] \cite{kozen99_autom}
  % Proof by reduction to the halting problem?
\end{thm}
\subsection{Incompleteness}
\begin{thm}[G\"odel's Incompleteness Theorem] 
  % Kleene 1943? Also shown in hopcroft 1979
  \cite{kleene43_recur_predic_quant}
\end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable.
\cite{sipser13:_introd_theor_comput}
\cite{post44:_recur}

@@

% \begin{defn}
%   A Pushdown Automata is...
% \end{defn}

@@

One of the earliest questions about the structure of the Turing degrees was
Post's Problem, first posed in 1944~\cite{post44:_recur}. It was solved
independently by Friedberg and Mucnik in the late 1950s using a technique
called the ``priority method''~\cite{Friedberg236, muchnik1956unsolvability}. A solution that does not
involve a priority argument was provided by Ku\v{c}era in 1986~\cite{Kucera:1986:APS:22416.22462}.

\begin{prob}[Post's Problem]  
  Does there exist a computably enumerable degree strictly between $0$ and $0'$?
  % Marchenkov 1976
  % Independently by Friedberg 57, Mucnik 56/58 using priority method
  % Priority free solution Kucer 1986''
  % ``Natural'' solns don't exist
\end{prob}

@@
